### 2016-06-29 testbed.tex
worked more on testbed description

After Raspbian lite has been unpacked, there should be an \texttt{img} file in the directory.
\texttt{fdisk} can be used to display the content of the image. Optionally, parse the arguments
"-u sectors" to display the sizes in sectors and "-l" to display the partitions within the
image and and automatically exit \texttt{fdisk} again:

@@

The output provides some useful knowledge about the image. First, the image is in total
2709504 sectors (1.3GiB) in size and contains two
partitions. The first partition starts at sector 8192 and
the other partition starts at sector 137216. The partitions is of type FAT32
and Linux and have sizes 63M and 1.2G. This indicates that the first partition
is most likely a boot partition and the second partition is a traditional Linux file
system. Actually, it is the Linux root filesystem (i.e. /).

\subsection{Resize image}
1.2G is not very much to hold the root filesystem given that it is desired to
in later steps alter the files and install more packages. It may therefore be
necessary to increase the partition size slightly.
The following steps illustrates how the image and its root filesystem
can be expanded by one \ac{GB}.

@@

Use \texttt{fdisk} to check that the image is one \ac{GB} bigger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
We use \texttt{fdisk} to alter the partition table. The echo's within the parenthesis
will function as
key-presses in \texttt{fdisk} and hence delete last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

@@

It is seen that the image is now available as \texttt{/dev/loop0}. Use \texttt{lsblk}
to view the partitions:

@@

Lets first mount the root partition. This is done by creating an empty directory
that we will be used as a mountpoint. We name it \texttt{root} and create it in the
working directory.
After this, simply mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

@@

After this, we can also mount the boot partition inside the newly mounted
root filesystem. This is convenient because that partition will be mounted
on this directory when a \ac{Raspi} starts up with a memory card containing
the image. %It is therefore the natural place to mount it for later purposes.

Mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

We can now change all the files in the disk image as desired.

>>>>

After Raspbian lite has been unpacked, there should be an \texttt{img} file in the directory.
\texttt{fdisk} can be used to display the content of the image. Parse the arguments
"-u sectors" to display the sizes in sectors and "-l" to display the partitions within the
image and and automatically exit \texttt{fdisk} again:

@@

The output provides knowledge about the image. The image is in total
2709504 sectors (1.3GiB) in size and contains two
partitions. The first partition starts at sector 8192 and
the other partition starts at sector 137216. The partitions is of type FAT32
and Linux and are 63M and 1.2G big. This indicates that the first partition
is most likely a boot partition and the second partition is a traditional Linux file
system. Actually, it is the Linux root filesystem (i.e. /).

\subsection{Resize image}
1.2G is not very much to store the root filesystem given that it is desired to
alter the files and install additional packages. It may therefore be
necessary to increase the partition size.
The following illustrates how the image and its root filesystem
can be expanded by one \ac{GB}.

@@

Use \texttt{fdisk} to see that the image is now one \ac{GB} larger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
Use \texttt{fdisk} to alter the partition table. The \texttt{echo}'s within the parenthesis
will be interpreted as
key-presses in \texttt{fdisk} and hence delete the last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

@@

\texttt{lsblk} may be used to list block devices. It appears that the image is
available as \texttt{/dev/loop0} which has two partitions \texttt{loop0p1} and
\texttt{loop0p2}:
%Use \texttt{lsblk} to view the partitions:

@@

Lets first mount the root partition. This is done by creating an empty directory
that we will be used as a mountpoint. We name it \texttt{root} and create it in the
working directory before mounting the root filesystem onto the mountpoint.
%After this, mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

@@

The root filesystem already has a boot directory that can be used as
the mountpoint for the boot partition.
%After this, we can also mount the boot partition inside the newly mounted
%root filesystem.
This is convenient because that partition will be mounted
on this directory when a \ac{Raspi} starts up with a memory card containing
the image. %It is therefore the natural place to mount it for later purposes.
Now, mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

%We can now change all the files in the disk image as desired.
It is now possible to change all files within the Raspbian image as desired.
We will take advantage of this to edit configuration files, append new files,
and even update and install packages.
